Separating Use and Reuse to Improve Both

### 2017-03-31 FACS/Examples.tex
.

To declare a trait \mTD\ or a class \mCD, we can use either a code literal \mL\ or a trait
expression.  Traits come with various operators (restrict, hide,
alias) but for now we focus on the single operator 
$\use$, taking a set
of code values: that is trait names \mt\ or literals \mL\ and composing them in an \emph{associative}
and \emph{commutative} way.  This operation, sometimes called \emph{sum}, is the simplest and most elegant
composition operator.  $\use\ \overline\mV$ composes the content of $\overline\mV$
by taking the union of the methods and the union of the implementations.

Use can not be applied if multiple versions of the same method are
present in different traits.  An exception is done for abstract methods:
methods where the implementation \me\ is missing; In this case (if the
headers are compatible) the implemented version is selected.  In a sum
of two abstract methods with compatible headers, the one with the more
specific type is selected. We will discuss later the formal details of
Sum.\marco{actually, this makes the sum not strongly associative...}

@@

Note how since our only point of relaxation is the class literal, there is no way an error can ``move around'' and be duplicated during the compilation process.
In particular, our approach allows for safe libraries of traits and classes to be fully typechecked, deployed and reused by multiple clients: no type error will emerge from library code.
On the other side, we do not enforce the programmer to write always self-contained code where all the abstract method definition are explicitly declared.

\section{Managing State}

The Sum operator is very elegant, but our research community is struggling to
make it work with constructor/state/fields.
The goals fo this struggle are as follow:

\begin{itemize}

\item keep sum associative and commutative.
\item allowing a class to create instances of itself.
\item actually initialize objects, not leaving no null fields
\item managing fields in a way that borrows the elengance of summing methods
\item make easy to add new fields

\end{itemize}

@@

  Pro: This approach is associative and commutative, even self construction
  can be allowed if the trait requires a static/class method
  returning This; the class will then implement this method by forwarding
  a call to the constructor.
  Negative: writing the class code with the constructors and
   fields and getter/setters and factories can be quite tedious.
   There is no way for a trait to specify a default value for a field,
   the class need to handle all the state, even state that is conceptually
   "private" of such trait.

  for example

\begin{lstlisting}
  pointSum: { method int x(); method int y();
    static method This of(int x,int y);
    method This sum(This other)
      This.of(this.x+other.x,this.y+other.y);
    }
  pointMul: { method int x(); method int y();
    static method This of(int x,int y);
    method This mul(This other)
      This.of(this.x*other.x,this.y*other.y);
    }
\end{lstlisting}

\subsection{A First Attempt at Composition}

\begin{lstlisting}
  Point:Use pointSum,pointMul
\end{lstlisting}

\noindent would fail since methods x,y and of are still abstract.
In this model the user is required to write something similar to

\begin{lstlisting}
  CPoint:Use point,colored, {
    int x; int y;
    method int x()x; method int y()y;
    class method This of(int x, int y)
      new Point(x,y);
    constructor Point(int x, int y){
      this.x=x;this.y=y;this.color=color;
      }
    }
\end{lstlisting}

\noindent after a while programming in this style, those "fixpoint" close the state in
the obvious way classes becomes quite repetitive, and one wonder
if it could be possible to automatically generate such code[]
In our model we go one step further:
In our model there is no need to generate code, or to explicitly
write down constructors and fields; there is not even syntax for those
constructor.
The idea is that any class that "could" be completed in the obvious way
is a complete "coherent" class.
Others call abstract a class that have abstract methods.
We call abstract a class that have a set of abstract methods that are not
coherent, that is, you can not interpret those as factory,getters and setters.

In the detail: a class with no abstract method is coherent, and like Java Math
will just be usefull for calling class/static methods.
a class with a single abstract class method returning This
is coherent if all the other abstract methods can be seen as "abstract state
operations" over one of its argument.
For example,
if there is a class method This of(int x, int y) as before,
then
a method int x() is an abstract state method, getter for x.
a method Void x(int that) is abstract state method, setter for x.
while getters and setters are fundamental operations, we can immagine
more operations to be supported:

\begin{lstlisting}
method This withX(int that) 
\end{lstlisting}
may be a "wither", doing a functional field update
  method Void update(int x,int) may do two field update at a time
  method This clone() may do a shallow clone of the object.
  We are not sure what is the best set of abstract state operations yet.

@@

\begin{lstlisting}
  pointSum: { method int x(); method int y();
    static method This of(int x,int y);
    method This sum(This other)
      This.of(this.x+other.x,this.y+other.y);
    }
  pointMul: { method int x(); method int y();
    static method This of(int x,int y);
    method This mul(This other)
      This.of(this.x*other.x,this.y*other.y);
    }
  PointAlgebra:Sum pointSum,pointMul
\end{lstlisting}

  As you can see, we can declare the methods independently and compose the result
  as we wish. However we have to repeat the abstract methods x,y and of.
  In addition of Sum,Mul we may want many operations over points; can we improve our reuse
  and not repeat such abstract definitions? of course!

\begin{lstlisting}
  p: { method int x(); method int y();
    static method This of(int x,int y);
    }
  pointSum:Sum p, { method This sum(This other)
      This.of(this.x+other.x,this.y+other.y);
    }
  pointMul:Sum p, { method This mul(This other)
      This.of(this.x*other.x,this.y*other.y);
    }
  pointDiv: ...
  PointAlgebra:Sum pointSum,pointMul,pointDiv,...
\end{lstlisting}

now our code is fully modularized, and each trait handle exactly one method.

@@

\begin{lstlisting}
  colored:{ method Color color() }
  Point:Sum pointSum,colored
\end{lstlisting}

This first attempt does not work: the abstract color method
is not a getter for any of the parameters of

\begin{lstlisting}
  class method This of(int x,int y) 
\end{lstlisting}

A solution is to provide a richer factory:

@@

where we assume to support overloading on different parameter number.
This is a good solution, we think is better that any alternatives in literature,
however the method CPoint.sum reset the color to red.
What should be the behaviour in this case?
If instead of writing This.of() we used 
this.withX(newX).withY(newY) we could preserve the color from this.
Not sure if that would be better.
If the point designer could predict this kind of extension, then we could
use the following design:  
\begin{lstlisting}
  p: { method int x(); method int y();
    method This withX(int that);
    method This withY(int that);
    static method This of(int x,int y);
    method This merge(This other);
    }
  pointSum:Sum p, { method This sum(This other)
      this.merge(other).withX(this.x+other.x).withY(this.y+other.y);
    }
  colored:{method Color color();
    method This withColor(Color that);
    method This merge(This other)
      this.withColor(this.color().mix(that.color());
    }
  CPoint:/*as before*/
\end{lstlisting}  
  Now we can merge colors, or any other kind of state we may want to add
  following this pattern.

>>>>

To declare a trait \mTD\ or a class \mCD, we can use either a code literal \mL\ or a trait
expression.  Traits come with various operators (restrict, hide,
alias) but for now we focus on the single operator 
$\use$, taking a set
of code values: that is trait names \mt\ or literals \mL\ and composing them in an \emph{associative}
and \emph{commutative} way.  This operation, sometimes called \emph{sum}, is the simplest and most elegant
composition operator.  $\use\ \overline\mV$ composes the content of $\overline\mV$
by taking the union of the methods and the union of the implementations.
In our simple mode, we consider an error trying to merge a class and an interface.

Use can not be applied if multiple versions of the same method are
present in different traits.  An exception is done for abstract methods:
methods where the implementation \me\ is missing; In this case (if the
headers are compatible) the implemented version is selected.  In a sum
of two abstract methods with compatible headers, the one with the more
specific type is selected. We will discuss later the formal details of
\use.\marco{actually, this makes the sum not strongly associative...}

@@

Note that we can reuse code only by naming traits; but our only point of relaxation is the class literal: there is no way an error can ``move around'' and be duplicated during the compilation process.
In particular, our approach allows for safe libraries of traits and classes to be fully typechecked, deployed and reused by multiple clients: no type error will emerge from library code.
On the other side, we do not enforce the programmer to write always self-contained code where all the abstract method definition are explicitly declared.

\section{Managing State}
This idea of summing pieces of code is very elegant,
and has proven very successfull in module composition languages~\cite{} but our research community is struggling to
make it work with object state (constructor and fields) while acheving the following goals:

\begin{itemize}
\item keep sum associative and commutative,
\item allowing a class to create instances of itself,
\item actually initialize objects, leaving no null fields,
\item managing fields in a way that borrows the elengance of summing methods,
\item make easy to add new fields.
\end{itemize}

@@

\paragraph*{Advantages:} 
this approach is associative and commutative, even self construction
  can be allowed if the trait requires a static/class method
  returning This; the class will then implement this method by forwarding
  a call to the constructor.

\paragraph*{Negatives:} writing the class code with the constructors and
   fields and getter/setters and factories can be quite tedious.
   Moreover, there is no way for a trait to specify a default value for a field,
   the class need to handle all the state, even state that is conceptually
   "private" of such trait.

Here in the following an exaple of such approach:

\begin{lstlisting}
  pointSum: { method int x()  method int y()
    class method This of(int x,int y)
    method This sum(This other)
      This.of(this.x+other.x,this.y+other.y)
    }
  pointMul: { method int x() method int y()
    class method This of(int x,int y)
    method This mul(This other)
      This.of(this.x*other.x,this.y*other.y)
    }
\end{lstlisting}

As you see, all the state operations are represented as abstract methods.

\subsection{A First Attempt at Composition}
According to the general ideas expressed before,
\begin{lstlisting}
  Point:Use pointSum,pointMul
\end{lstlisting}

\noindent would fail since methods \Q@x@,\Q@y@ and \Q@of@ are still abstract.
In this mindset, the user would be required to write something similar to

\begin{lstlisting}
  CPoint:Use point,colored, {//not our suggested solution
    int x; int y;
    method int x()x; method int y()y;
    class method This of(int x, int y)
      new Point(x,y);
    constructor Point(int x, int y){
      this.x=x;this.y=y;this.color=color;
      }
    }
\end{lstlisting}

\noindent after a while programming in this style, 
writing those obvious ``close the state'' classes become a
repetivie boring job, and one wonder
if it could be possible to automatically generate such code[].
Indeed those classes are just a form of ``fixpoint''.

In our model we go one step further: there is no need to generate code, or to explicitly
write down constructors and fields; there is not even syntax for those
constructor.
The idea is that any class that ``could" be completed in the obvious way
\emph{is a complete ``coherent" class}.
In most other languages, a class is abstract if have abstract methods.
Instead, we call abstract a class whose set of abstract methods is not
coherent, that is, can not be automatically recognized as factory,getters and setters.

\paragraph*{Detaild definition of coherent:}
\begin{itemize}
\item a class with no abstract method is coherent, and like Java \Q@Math@.
Will just be usefull for calling class/static methods.
\item a class with a single abstract \Q@class@ method returning \Q@This@
is coherent if all the other abstract methods can be seen as \emph{abstract state
operations} over one of its argument.
For example,
if there is a \Q@class method This of(int x, int y)@ as before,
then
\item a method \Q@int x()@ is intepreted as an abstract state method: a getter for \Q@x@.
\item a method \Q@Void x(int that)@ is a setter for x.
\end{itemize}

While getters and setters are fundamental operations, we can immagine
more operations to be supported; for example
\begin{itemize}
\item \Q@method This withX(int that)@
may be a ``wither", doing a functional field update.
\item \Q@method Void update(int x,int)@
may do two field update at a time.
\item\Q@method This clone()@ may do a shallow clone of the object.
\end{itemize}

We are not sure what is the best set of abstract state operations yet, but we think this could become a very interesting area of research.

@@

\begin{lstlisting}
 //same code as before for pointSum and pointMul
  pointSum: { method int x() method int y()
    static method This of(int x,int y)
    method This sum(This other)
      This.of(this.x+other.x,this.y+other.y)
    }
  pointMul: { method int x() method int y()//look we are repeating
    static method This of(int x,int y)//the abstract method declarations.
    method This mul(This other)
      This.of(this.x*other.x,this.y*other.y)
    }
  PointAlgebra:Sum pointSum,pointMul
\end{lstlisting}

  As you can see, we can declare the methods independently and compose the result
  as we wish. However we have to repeat the abstract methods \Q@x@,\Q@y@ and \Q@of@.
  In addition of \Q@Sum@ and \Q@Mul@ we may want many operations over points; can we improve our reuse
  and not repeat such abstract definitions? of course!

\begin{lstlisting}
  p: { method int x() method int y()
    static method This of(int x,int y)
    }
  pointSum:Use p, { method This sum(This other)
      This.of(this.x+other.x,this.y+other.y)
    }
  pointMul:Use p, { method This mul(This other)
      This.of(this.x*other.x,this.y*other.y)
    }
  pointDiv: ...
  PointAlgebra:Use pointSum,pointMul,pointDiv,...
\end{lstlisting}

Now our code is fully modularized, and each trait handle exactly one method.

@@

\begin{lstlisting}
  colored:{ method Color color() }
  Point:Sum pointSum,colored //fails
\end{lstlisting}

This first attempt does not work: the abstract color method
is not a getter for any of the parameters of 
\Q@ class method This of(int x,int y)@
A solution is to provide a richer factory:

@@

where we assume to support overloading on different parameter number.
This is a good solution, we think is better that any alternatives in literature,
however the method CPoint.sum reset the color to red.
What should be the behaviour in this case?
If we support withers, instead of writing This.of() we can use
this.withX(newX).withY(newY) in order to preserve the color from \Q@this@.
Sadly, if we use this design inside of \Q@sum(This other)@ we would loose the color from \Q@other@.

If the point designer could predict this kind of extension, then we could use the following design:  
\begin{lstlisting}
  p: { method int x() method int y()
    method This withX(int that)
    method This withY(int that)
    static method This of(int x,int y)
    method This merge(This other)
    }
  pointSum:Use p, { method This sum(This other)
      this.merge(other).withX(this.x+other.x).withY(this.y+other.y);
    }
  colored:{method Color color()
    method This withColor(Color that)
    method This merge(This other)
      this.withColor(this.color().mix(that.color());
    }
  CPoint:/*as before*/
\end{lstlisting}  
  Now we can merge colors, or any other kind of state we may want to add
  following this pattern.
  In order to compose, let say \Q@colored@ with \Q@flavored@ we would
  need to compose the merge operation inside of both of them.
  The simple model we are presenting could accomodate this with an
  extension allowing code literals inside of a \use\ expression to use some form of super call to compose conflicting implementations. This is similar to the \emph{override} operation present in the original trait model[].
